Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-05-09 text/Manuscript.Rmd

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates *soluble guanylate cyclase (sGC)* expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii* **(Figure 1A)**, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses *c-opsin1*, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in *c-opsin1* knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017]. Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs **(Figure 4---figure supplement 1)**. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and 2 choanoflagellate species. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 *Platynereis* sequences were parts of several major and ancient clades of the tree **(Figure 4---figure supplement 1)**. To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells **(Figure 4---figure supplement 1)**. In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers **(Figure 4---figure supplement 2A)**. The INNOS cells were identified by NOS expression and their spatial mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out in situ HCR with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *MLD proneuropeptide* **(Figure 4A,B and Figure 4---figure supplement 2A-C)**. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus **(Figure 4C)**. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus **(Figure 4---figure supplement 2D)**. In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs **(Figure 4D)**. These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

>>>>

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neursal circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates *soluble guanylate cyclase (sGC)* expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii* **(Figure 1A)**, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses *c-opsin1*, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in *c-opsin1* knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017]. Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC)[@Shu2003; @monteagudo-cascales2023] as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs **(Figure 4---figure supplement 1)**. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and 2 choanoflagellate species. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 *Platynereis* sequences were parts of several major and ancient clades of the tree **(Figure 4---figure supplement 1)**. To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells **(Figure 4---figure supplement 1)**. In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers **(Figure 4---figure supplement 2A)**. The INNOS cells were identified by NOS expression and their spatial mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out in situ HCR with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *MLD proneuropeptide* **(Figure 4A,B and Figure 4---figure supplement 2A-C)**. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus **(Figure 4C)**. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus **(Figure 4---figure supplement 2D)**. In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs **(Figure 4D)**. These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

'''{r cars}

readxl
